Finally Tagless, Partially Evaluated

### 2008-09-12 tagless/tagless-final.tex
Revised Section 4

To reiterate, we do \emph{not} propose any new language
feature or \ifshort new \else even any new programming \fi technique.
\ifshort
We
\else
Rather, we
\fi
solve a problem that was stated in the published record as open and
likely unsolvable in ML or Haskell 98 without extensions, by a novel
combination of simple types and techniques already described in the
literature that use features present in mainstream functional languages.
In particular, we follow \citets{yang-encoding} encoding of type-indexed
values, \citets{Sperber-SelfApplicable} and \citets{asai-binding-time}
construction of dynamic terms alongside static terms, and
\citets{sumii-hybrid} deforestation of syntax
constructors.  These techniques require just
a Hindley-Milner type system with either module
functors or constructor classes, as realized in all variants of ML and
Haskell.
The simplicity of our solution and its
use of only mainstream features \ifshort\else are virtues that \fi make it more practical to build typed,
embedded DSLs.

@@

\section{A tagless partial evaluator}\label{PE}
Surprisingly, the |Symantics| interface presented above extends to
encompass an online partial evaluator.  We present this partial evaluator
in a sequence of three attempts. It uses no universal type and no tags
for object types.  We then discuss residualization and binding-time
analysis.  Our partial evaluator is a modular extension of the evaluator
in~\S\ref{S:interpreter-RL} and the compiler in~\S\ref{S:compiler}, in
that it uses the former to reduce static terms and the latter to build
dynamic terms.

@@

Roughly, a partial evaluator interprets each object term to yield either
a static (present-stage) term (using the evaluator~|R|) or
a dynamic (future-stage) term (using the compiler~|C|).  To
distinguish between static and dynamic terms, we might try to define
|repr| in the partial evaluator (module |P0|) as
\begin{code}
type ('c,'dv) repr = S0 of ('c,'dv) R.repr | D0 of ('c,'dv) C.repr
\end{code}
(the digit zero in the names of the phase tags |S0| and |D0| is the
version suffix).
To extract a dynamic term from this type, we create the function
\begin{code}
let abstrI0 (e:('c,int) repr) : ('c,int) C.repr =
  match e with
  | S0 e -> C.int e
  | D0 e -> e
\end{code}
and the similar function |abstrB0| for dynamic boolean terms. Here, 
|C.int| is used to convert from the static term |('c,int) R.repr|, 
which is just |int|, to a dynamic term. We can now define the following
components of module |P0| required by the |Semantics| signature:
\begin{code}
let int  (x:int)  = S0 (R.int x)
let bool (x:bool) = S0 (R.bool x)
let add e1 e2 = 
  match (e1,e2) with
  | (S0 e1, S0 e2) -> S0 (R.add e1 e2)
  | (e1, e2)       -> D0 (C.add (abstrI0 e1) (abstrI0 e2))
\end{code}
Integer and boolean literals are immediate, present-stage
values. Addition yields a static term (using~|R.add|) if and only 
if both operands are static; otherwise we extract the dynamic terms 
from the operands and add them using~|C.add|.

@@

Our |Symantics| only includes separate lifting methods |bool| and
|int|, not a parametrically polymorphic lifting method, for good reason:
When compiling to a first-order target language such as machine code,
booleans, integers, and functions may well be represented differently.
Compiling a polymorphic lift function thus requires intensional type
analysis.  To avoid needing polymorphic lift, we turn to
\citets{Sperber-SelfApplicable} and \citets{asai-binding-time} technique of
building a dynamic term alongside every static term \citep{sumii-hybrid}.

@@

We start building the partial evaluator anew (module |P1|) and
switch to the data type
\begin{code}
type ('c,'dv) repr = P1 of ('c,'dv) R.repr option * ('c,'dv) C.repr
\end{code}
so that a partially evaluated term always contains a dynamic
component and sometimes contains a static component.  By 
distributivity, the two
alternative constructors of an |option| value, |Some| and |None|,
tag each partially evaluated term with a phase: either present or
future.  This tag is not an object type tag: all pattern matching below
is exhaustive. Because the future-stage component is always available, we
can now define the polymorphic function
|let abstr1 (P1 (_,dyn)) = dyn| of type
\texttt{('c,'dv) repr -> ('c,'dv) C.repr}
to extract it without requiring polymorphic lift into~|C|.  We then try
to define the interpreter |P1|---and get as far as the first-order
constructs of our object language, including |if_|.
\begin{code}
module P1 : Symantics = struct
  let int (x:int) = P1 (Some (R.int x), C.int x)
  let add e1 e2 = match (e1,e2) with
    | (P1 (Some n1,_),P1 (Some n2,_)) -> int (R.add n1 n2)
    | _ -> P1 (None,(C.add (abstr1 e1) (abstr1 e2)))
  let if_ = function
    | P1 (Some s,_) -> fun et ee -> if s then et () else ee ()
    | eb -> fun et ee -> P1 (None, C.if_ (abstr1 eb) 
                                   (fun () -> abstr1 (et ()))
                                   (fun () -> abstr1 (ee ())))
\end{code}
However, we stumble on functions.  According to our
definition of~|P1|, a partially evaluated object function, such as the
identity $\fun{x}x$ embedded in OCaml as |lam (fun x -> x)|\texttt{ :
}|('c,'a->'a) P1.repr|, consists of a dynamic part 
(type |('c,'a->'a) C.repr|) and
maybe a static part (type |('c,'a->'a) R.repr|).  The dynamic part is useful
when this function is passed to another function that is only
dynamically known, as in $\fun{k}k(\fun{x}x)$.  The static part is
useful when this function is applied to a static argument, as in
$(\fun{x}x)\True$.  Neither part, however, lets us \emph{partially}
evaluate the function, that is, compute as much as possible statically
when it is applied to a mix of static and dynamic inputs.  For example,
the partial evaluator should turn $\fun{n}(\fun{x}x)n$ into $\fun{n}n$
by substituting $n$ for~$x$ in the body of $\fun{x}x$ even though $n$ is
not statically known.  The same static function, applied to
different static arguments, can give both static and dynamic results: we
want to simplify $(\fun{y}x\times y)0$ to~$0$ but $(\fun{y}x\times y)1$
to~$x$.

To enable these simplifications, we delay binding-time analysis
for a static function until it is applied, that is, until |lam f|
appears as the argument of |app|.  To do so, we have to incorporate |f|
as it is into the |P1.repr| data structure: the representation
for a function type |'a->'b| should be one of
\begin{code}
S1 of ('c,'a) repr -> ('c,'b) repr | E1 of ('c,'a->'b) C.repr
P1 of (('c,'a) repr -> ('c,'b) repr) option * ('c,'a->'b) C.repr
\end{code}
unlike |P1.repr| of |int| or |bool|.
That is, we need a nonparametric data type, something akin to
type-indexed functions and type-indexed types, which
\citet{oliveira-typecase} dub the \emph{typecase} design pattern.
Thus, typed partial evaluation, like typed CPS transformation
(see \S\ref{S:CPS}),
inductively defines a map from source types to target types that
performs case distinction on the source type. In Haskell, typecase
can be implemented using either GADTs or
type-class functional dependencies
\citep{oliveira-typecase}. The accompanying code shows both
approaches (|Incope.hs| and |incope1.hs|), 
neither of which is portable to OCaml. In addition,
the problem of non\hyp exhaustive pattern\hyp matching reappears in
the GADT approach because GHC 6.8 and prior cannot see that a particular
type of GADT value precludes certain constructors. Although this
is an implementation issue of GHC,
it indicates that assuring exhaustive pattern match with GADTs
requires non-trivial reasoning (beyond the abilities of GHC at the moment);
certainly GADTs fail to
make it \emph{syntactically} apparent that pattern matching is exhaustive.

@@

Let us re-examine the problem in~\S\ref{S:PE-problem}. What we
would ideally like is to write
\begin{code}
type ('c,'dv) repr = P1 of (repr_pe ('c,'dv)) R.repr option * ('c,'dv) C.repr
\end{code}
where |repr_pe| is the type function defined
% inductively because P below depends on repr_pe
by
\begin{code}
repr_pe ('c,int)    = ('c,int)
repr_pe ('c,bool)   = ('c,bool)
repr_pe ('c,'a->'b) = ('c,'a) repr -> ('c,'b) repr
\end{code}
Although we can use type classes to define this type function
in Haskell, that is not portable to MetaOCaml. However,
these three typecase alternatives are already present in existing
methods of |Symantics|.
A simple and portable solution\footnote{though somewhat long-winded} thus
emerges: we bake |repr_pe| into the signature |Symantics|. 
We recall from Figure~\ref{fig:ocaml-simple} in~\S\ref{encoding} that 
the |repr| type constructor took two arguments |'c| and~|'dv|.
We add an argument |'sv| for the result of applying |repr_pe| to~|'dv|.
Figure~\ref{fig:ocaml} shows the new signature.
\begin{figure}
\begin{floatrule}
\begin{code2}
module type Symantics = sig
  type ('c,'sv,'dv) repr

@@

The interpreters |R|, |L| and~|C| in \S\ref{S:interpreter-RL},\S\ref{S:compiler}
only use the old
type arguments |'c| and~|'dv|, which are treated by the new signature
in the same way.  Hence, all that needs to change in these interpreters
to match the new signature is to add a phantom type
argument~|'sv| to~|repr|.
For example, the compiler |C| now begins
\begin{code}
module C = struct
  type ('c,'sv,'dv) repr = ('c,'dv) code
\end{code}
with the rest the same.
Predictably, the partial evaluator~|P| crucially relies on the type argument
|'sv|.

Figure~\ref{fig:pe} shows the partial evaluator~|P|.
Its type |repr| expresses the type definition for |repr_pe| given
at the start of this section, with |'sv| taking the place of |repr_pe t|.
The function |abstr|
extracts a future-stage code value from the result of
partial evaluation.  Conversely, the function |pdyn| injects a
code value into the |repr| type. As
in~\S\ref{S:PE-problem}, we build dynamic terms alongside
any static ones to express how the lift function is indexed
by the dynamic type.  What is new is that we analogously
build a static type alongside the dynamic type to express
how the static type is indexed by the dynamic type.

The static portion of the interpretation of |lam f| is |Some f|, 
which just wraps the HOAS
function |f|. The interpretation of |app ef ea| 
checks to see if |ef| is such a wrapped
HOAS function. If it is, we apply |f| to the
concrete argument |ea|, giving us a chance to perform static
computations (see the example below). If
|ef| has only a dynamic part, we residualize.

To illustrate how to add optimizations, we improve |add| (and |mul|,
elided) to simplify the generated code using the monoid (and ring)
structure of~|int|: not only is addition performed statically
(using~|R|) when both operands are statically known, but it is
eliminated when one operand is statically~$0$; similarly for
multiplication by~$0$ or~$1$.  
Such algebraic simplifications are easy
to abstract over the specific domain (such as monoid or ring) where they
apply.  These simplifications and abstractions help a lot
in a large language with more base types and primitive operations.
Incidentally, the accompanying code actually contains a more general
implementation mechanism for such features, inspired in part by previous work
in generative linear algebra~\citep{CaretteKiselyov05}.

@@

\noindent All pattern\hyp
matching in~|P| is \emph{syntactically} exhaustive, so it is patent to the
metalanguage implementation that |P| never gets stuck.  Further, all
pattern\hyp matching occurs
during partial evaluation, only to check if a value is known statically,
never what type it has.  In other words, our partial evaluator tags
phases (with |Some| and |None|) but not object types.

Our typed partial evaluator is online and polyvariant.  It reuses the
compiler~|C| and the evaluator~|R| by composing them.  This situation is
simpler than \citets{SperberThiemann:TwoForOne} composition of a partial
evaluator and a compiler, but the general ideas are similar.
\fi

@@

We would like to collect those interpretation functions into a module
with signature |Symantics|, to include the CBN CPS interpreter within our
general framework. Alas, as in~\S\ref{S:PE-problem}, the type of
an object term inductively determines the type of its interpretation:
the interpretation of an object term of type~$t$ may not have type
|(|$t$|->'w)->'w|, because $t$ may be a function type.  Again we
simulate a type function with a typecase distinction, using an extra
type argument to |repr|. Luckily, the type function |repr_pe| needed for
the partial evaluator 
in~\S\ref{S:PE-solution} is precisely the same type function we
need for CBN CPS\@.
\begin{code}
module RCN = struct
  type ('c,'sv,'dv) repr = {ko: 'w. ('sv -> 'w) -> 'w}
  let int (x:int) = {ko = fun k -> k x}
  let add e1 e2 = {ko = fun k ->
      e1.ko (fun v1 -> e2.ko (fun v2 -> k (v1 + v2)))}
  let if_ eb et ee = {ko = fun k ->
      eb.ko (fun vb -> if vb then (et ()).ko k else (ee ()).ko k)}
  let lam f = {ko = fun k -> k f}
  let app e1 e2 = {ko = fun k -> e1.ko (fun f -> (f e2).ko k)}
  let fix f = let rec fx f n = app (f (lam (fx f))) n in lam (fx f)
  let run x = x.ko (fun v -> v)
end
\end{code}

@@

The module returned by |CPST| does not define |repr|
and thus does not have signature |Symantics|.
The reason is again the type of |lam f|. Whereas
|int| and |add| return the (abbreviated) type
|('c, ..., (int -> 'w) -> 'w) S.repr|,
the type of \texttt{lam (add (int~1))} is
\begin{code}
('c, ..., ((int -> (int -> 'w1) -> 'w1) -> 'w2) -> 'w2) S.repr
\end{code}
Hence, to write the type equation defining |CPST.repr| we again need
a type function with a typecase distinction, similar to |repr_pe|
in~\S\ref{S:PE-solution}. Alas, the type function we need is not
identical to |repr_pe|, so we need to add another type argument to
|repr| in the |Symantics| signature. As in~\S\ref{S:PE-solution}, the
terms in previous implementations of |Symantics| stay unchanged, but the
|repr| type equations in those implementations have to take a new
(phantom) type argument.
The verbosity of these types is the only difficulty in defining a
new signature to replace |Symantics| that the output of |CPST| matches.

@@

Our partial evaluator establishes a bijection |repr_pe| between static
and dynamic types (the valid values of |'sv| and |'dv|), and between
static and dynamic terms.  It is customary to implement such a bijection
using an injection\hyp projection pair, as done for interpreters
\ifshort \citep{Ramsey-ML-module-mania,Benton-embedded-interpreters}\else
by \citet{Ramsey-ML-module-mania} and \citet{Benton-embedded-interpreters}\fi,
partial evaluation \ifshort \citep{Danvy-TDPE}\else by \citet{Danvy-TDPE}\fi,
and type-level functions \ifshort \citep{oliveira-typecase}\else by
\citet{oliveira-typecase}\fi.  As explained in~\S\ref{S:PE-solution}, we
avoid injection and projection at the type level by adding an argument
to |repr|.
Our solution could have been even more straightforward if MetaOCaml
provided total type-level functions such as |repr_pe| in
\S\ref{S:PE-solution}---simple type-level computations 
ought to become mainstream.

At the term level, we also avoid converting between static and dynamic
terms by building them in parallel, as \citet{Sperber-SelfApplicable}
and \citet{asai-binding-time} do in their partial evaluators.
This method type-checks in Hindley-Milner once we
deforest the object term representation.  Put another way, we
manually apply type-level partial evaluation to our type
functions (see \S\ref{S:PE-solution}) to obtain simpler types 
acceptable to MetaOCaml.

\Citets{JFP-Mogensen} self-reducer for the untyped
$\lambda$-calculus preceded Sperber's and Asai's partial evaluators in constructing
static and dynamic terms in parallel.  However, this self-reducer
constructs a static term for every object term, even a bound variable,
so (in terms of Fig.~\ref{fig:pe}) it moves some work from |app|
to~|pdyn|.  In contrast, we follow Sperber and Asai in constructing the static term
optionally, so as to perform type-directed lifting without requiring the
encoding of an object term to juggle explicit type indices.

\Citet{sumii-hybrid} also use Sperber and Asai's construction,
to combine online
and offline partial evaluation.
They predate us in deforesting the object term representation to enable tagless
partial evaluation.
We strive for modularity by reusing interpreters for individual stages
\citep{SperberThiemann:TwoForOne}: our partial evaluator~|P|
reuses our tagless evaluator~|R| and tagless compiler~|C|, so it
is patent that the \emph{output} of~|P| never gets
stuck.  It would be interesting
to try to derive a \emph{cogen} \citep{Thiemann:cogeninsixlines}
in the same manner.

@@

\Citet{fiore:nbe-ppdp2002} and \citet{balat:tdpe-popl2004} also build
a tagless partial evaluator, using delimited control operators.  It is
type-directed, so the user must represent, as a term, the type of every
term to be partially evaluated.  We shift this work to the type checker
of the metalanguage.  By avoiding term-level type representations, our
approach makes it easier to perform algebraic simplifications (as
in~\S\ref{S:PE-solution}).

\oleg{The following are almost unedited excerpts from our
  correspondence with Sumii about Birkedal's thesis. We definitely
  should include this discussion and the reference to his thesis.}
\Citet{Birkedal-PE-ML} have hand-written a compiler generator (cogen)
for Core Standard ML, which transforms a program into its tagless
generating extension (and therefore serves as a tagless partial
evaluator, intepreter, and compiler in our sense). They still have to
use the "universal type" at the cogen level, which is removed from
compile-time, so that the end result is tagless indeed.  The main
difference between the work of \citet{Birkedal-PE-ML} and ours is how
\emph{early} it becomes \emph{manifest} to the metalanguage
that our code is safe.  The fact that
\citet{Birkedal-PE-ML}'s generating extension never generates a
specialized program that is ill-typed is not manifest to the SML
compiler until the SML compiler sees each particular specialized
program.  Moreover, and in a similar vein, the fact that
\citet{Birkedal-PE-ML}'s cogen never generates a generating extension
that is ill-typed (or that may generate a specialized program that is
ill-typed) is not manifest to the SML compiler until the SML compiler
sees each particular generating extension (or each particular
specialized program).  This early and manifest type safety is the
payoff of our treating an object program not as a data structure but
as calls to abstract-syntax constructor functions.  The tradeoff, as
we currently see it, is that we have to swear off intensional analysis
of code. \oleg{Unless we convert to our representation to GADT, which
  is always possible, if GADTs are available.}

@@

Our encoding of the type function |repr_pe| in \S\ref{S:PE-solution}
emulates type-indexed types and is related to intensional type analysis
\citep{Morrisett-intensional,Generic-Haskell}. However, our object
language and running examples in HOAS include |fix|,
which intensional type analysis cannot handle
\citep{xi-guarded}.

>>>>

To reiterate, we do \emph{not} propose any new language
feature or \ifshort new \else even any new programming \fi technique.
\ifshort
We
\else
Rather, we
\fi
solve a problem that was stated in the published record as open and
likely unsolvable in ML or Haskell 98 without extensions, by a novel
combination of simple types and techniques already described in the
literature that use features present in mainstream functional languages.
In particular, we follow \citets{yang-encoding} encoding of type-indexed
values, \citets{Sperber-SelfApplicable} and \citets{asai-binding-time}
construction of dynamic terms alongside static terms, and
\citets{Thiemann-combinators} deforestation of syntax
constructors.  These techniques require just
a Hindley-Milner type system with either module
functors or constructor classes, as realized in all variants of ML and
Haskell.
The simplicity of our solution and its
use of only mainstream features \ifshort\else are virtues that \fi make it more practical to build typed,
embedded DSLs.

@@

\section{A tagless partial evaluator}\label{PE}
Surprisingly, this |Symantics| interface extends to
encompass an online partial evaluator that uses no universal type and no tags
for object types.  We present this partial evaluator
in a sequence of three attempts to express the types
of residualization and binding-time
analysis.  Our partial evaluator is a modular extension of the evaluator
in~\S\ref{S:interpreter-RL} and the compiler in~\S\ref{S:compiler}, in
that it uses the former to reduce static terms and the latter to build
dynamic terms.

@@

Roughly, a partial evaluator interprets each object term to yield either
a static (present-stage) term (using the evaluator~|R|) or
a dynamic (future-stage) term (using the compiler~|C|).  To
distinguish between static and dynamic terms, we might try to define
|repr| in the partial evaluator as follows.
In the phase tags |S0| and~|D0|, the digit zero indicates
our initial attempt.
\begin{code}
type ('c,'dv) repr = S0 of ('c,'dv) R.repr | D0 of ('c,'dv) C.repr
\end{code}
To extract a dynamic term from this type, we create the function
\begin{code}
let abstrI0 (e : ('c,int) repr) : ('c,int) C.repr =
  match e with S0 e -> C.int e | D0 e -> e
\end{code}
and the similar function |abstrB0| for dynamic boolean terms. Here, 
|C.int| is used to convert a static term (of type |('c,int) R.repr|, 
which is just |int|) to a dynamic term. We can now define the following
components required by the |Semantics| signature:
\begin{code}
let int  (x:int)  = S0 (R.int x)
let bool (x:bool) = S0 (R.bool x)
let add e1 e2 = match (e1,e2) with
  | (S0 e1, S0 e2) -> S0 (R.add e1 e2)
  | _ -> D0 (C.add (abstrI0 e1) (abstrI0 e2))
\end{code}
Integer and boolean literals are immediate, present-stage
values. Addition yields a static term (using~|R.add|) if and only 
if both operands are static; otherwise we extract the dynamic terms 
from the operands and add them using~|C.add|.

@@

Our |Symantics| signature only includes separate lifting methods |bool| and
|int|, not a polymorphic lifting method, for good reason:
When compiling to a first-order target language such as machine code,
booleans, integers, and functions may well be represented differently.
Compiling a polymorphic lift function thus requires intensional type
analysis.  To avoid needing polymorphic lift, we turn to
\citets{Sperber-SelfApplicable} and \citets{asai-binding-time} technique of
building a dynamic term alongside every static term \citep{sumii-hybrid}.

@@

We start building the partial evaluator anew and
switch to the data type
\begin{code}
type ('c,'dv) repr = P1 of ('c,'dv) R.repr option * ('c,'dv) C.repr
\end{code}
so that a partially evaluated term always contains a dynamic
component and sometimes contains a static component.
The two
alternative constructors of an |option| value, |Some| and |None|,
tag each partially evaluated term to indicate whether its value is
known statically at the present stage.
This tag is not an object type tag: all pattern matching below
is exhaustive. Now that the future-stage component is always available, we
can define the polymorphic function
\begin{code}
let abstr1 (P1 (_,dyn) : ('c,'dv) repr) : ('c,'dv) C.repr = dyn
\end{code}
to extract it without needing polymorphic lift into~|C|.  We then try
to define the term combinators\nobreak\hspace{0pt}---and get as far as
the first-order constructs of our object language, including |if_|.
\begin{code}
let int (x:int) = P1 (Some (R.int x), C.int x)
let add e1 e2 = match (e1,e2) with
  | (P1 (Some n1, _), P1 (Some n2, _)) -> int (R.add n1 n2)
  | _ -> P1 (None, C.add (abstr1 e1) (abstr1 e2))
let if_ eb et ee = match eb with
  | P1 (Some s, _) -> if s then et () else ee ()
  | _ -> P1 (None, C.if_ (abstr1 eb) (fun () -> abstr1 (et ()))
                                     (fun () -> abstr1 (ee ())))
\end{code}
However, we stumble on functions.  Given how we just
defined~|repr|, a partially evaluated object function, such as the
identity $\fun{x}x$ (of type $\ZZ\to\ZZ$) embedded in OCaml as
|lam (fun x -> x)| (of type |('c,int->int) repr|), consists of
a dynamic part (of type |('c,int->int) C.repr|) and optionally
a static part (of type |('c,int->int) R.repr|).  The dynamic part is useful
when this function is passed to another function that is only
dynamically known, as in $\fun{k}k(\fun{x}x)$.  The static part is
useful when this function is applied to a static argument, as in
$(\fun{x}x)\True$.  Neither part, however, lets us \emph{partially}
evaluate the function, that is, compute as much as possible statically
when it is applied to a mix of static and dynamic inputs.  For example,
the partial evaluator should turn $\fun{n}(\fun{x}x)n$ into $\fun{n}n$
by substituting $n$ for~$x$ in the body of $\fun{x}x$ even though $n$ is
not statically known.  The same static function, applied to
different static arguments, can give both static and dynamic results: we
want to simplify $(\fun{y}x\times y)0$ to~$0$ but $(\fun{y}x\times y)1$
to~$x$.

To enable these simplifications, we delay binding-time analysis
for a static function until it is applied, that is, until |lam f|
appears as the argument of |app|.  To do so, we have to incorporate |f|
as is into |lam f|: the type |('c,'a->'b) repr| should be one of
\begin{code}
S1 of ('c,'a) repr -> ('c,'b) repr | E1 of ('c,'a->'b) C.repr
P1 of (('c,'a) repr -> ('c,'b) repr) option * ('c,'a->'b) C.repr
\end{code}
unlike |('c,int) repr| or |('c,bool) repr|.
That is, we need a nonparametric data type, something akin to
type-indexed functions and type-indexed types, which
\citet{oliveira-typecase} dub the \emph{typecase} design pattern.
Thus, typed partial evaluation, like typed CPS transformation
(see \S\ref{S:CPS}),
inductively defines a map from source types to target types that
performs case distinction on the source type. In Haskell, typecase
can be implemented using either GADTs or
type-class functional dependencies
\citep{oliveira-typecase}. The accompanying code shows both
approaches (|Incope.hs| and |incope1.hs|), 
neither of which is portable to OCaml. In addition,
the problem of non\hyp exhaustive pattern\hyp matching reappears in
the GADT approach because GHC 6.8 and prior cannot see that a particular
type of GADT value precludes certain constructors. Although this
is an implementation issue of GHC,
it indicates that assuring exhaustive pattern match with GADTs
requires non-trivial reasoning (beyond the abilities of GHC at the moment);
certainly GADTs fail to
make it \emph{syntactically} apparent that pattern matching is exhaustive.

@@

The problem in the last section is that we want to write
\begin{code}
type ('c,'dv) repr = P1 of ('c,'dv) static option * ('c,'dv) C.repr
\end{code}
where |static| is the type function defined
% inductively because P below depends on static
by
\begin{code}
('c,int)    static = ('c,int) R.repr
('c,bool)   static = ('c,bool) R.repr
('c,'a->'b) static = ('c,'a) repr -> ('c,'b) repr
\end{code}
Although we can use type classes to define this type function
in Haskell, that is not portable to OCaml. However,
the three typecase alternatives of |static| are already present in existing
methods of |Symantics|.
Thus emerges
a simple and portable solution, if a long-winded one:
we bake |static| into the signature |Symantics|. 
In Figure~\ref{fig:ocaml-simple},
the |repr| type constructor took two arguments |('c,'dv)|;
in Figure~\ref{fig:ocaml},
we add an argument |'sv| for the type |('c,'dv) static|.
\begin{figure}
\begin{floatrule}
\begin{code2}
module type Symantics = sig
  type ('c,'sv,'dv) repr

@@

The interpreters |R|, |L| and~|C| in \S\ref{S:interpreter-RL} and~\S\ref{S:compiler}
only use the old
type arguments |'c| and~|'dv|, which are treated by the new signature
in the same way.  Hence, all that needs to change in these interpreters
to match the new signature is to add a phantom type
argument~|'sv| to~|repr|.
For example, the compiler |C| now begins
\begin{code}
module C = struct
  type ('c,'sv,'dv) repr = ('c,'dv) code
\end{code}
with the rest the same.

Figure~\ref{fig:pe} shows the partial evaluator~|P|.
Its type |repr| expresses the definition for |static| given
at the start of this section, with |'sv| taking the crucial place of |('c,'dv) static|.
The function |abstr|
extracts a future-stage code value from the result of
partial evaluation.  Conversely, the function |pdyn| injects a
code value into the |repr| type. As
in~\S\ref{S:PE-problem}, we build dynamic terms alongside
any static ones to express how the lift function is indexed
by the dynamic type.  What is new is that we analogously
build a static type alongside the dynamic type to express
how the static type is indexed by the dynamic type.

The static portion of the interpretation of |lam f| is |Some f|, 
which just wraps the HOAS
function |f|. The interpretation of |app ef ea| 
checks to see if |ef| is such a wrapped
HOAS function. If it is, we apply |f| to the
concrete argument |ea|, so as to perform static
computations (see the example below). If
|ef| has only a dynamic part, we residualize.

To illustrate how to add optimizations, we improve |add| (and |mul|,
elided) to simplify the generated code using the monoid (and ring)
structure of~|int|: not only is addition performed statically
(using~|R|) when both operands are statically known, but it is
eliminated when one operand is statically~$0$; similarly for
multiplication by~$0$ or~$1$.  
Although our basic machinery for partial evaluation is independent of
such algebraic simplifications, it makes them easy to add and
to abstract over the specific domains (such as monoid or ring) where they
apply.  These simplifications and abstractions help a lot
in a large language with more base types and primitive operations.
Incidentally, the accompanying code actually contains a more general
implementation mechanism for such features, inspired in part by previous work
in generative linear algebra~\citep{CaretteKiselyov05}.

@@

All pattern\hyp
matching in~|P| is \emph{syntactically} exhaustive, so it is patent to the
metalanguage implementation that |P| never gets stuck.  Further,
|P| uses pattern\hyp matching
only to check if a value is known statically,
never to check what type a value has dynamically.
In other words, our partial evaluator tags
phases (with |Some| and |None|) but not object types, so it
is patent that the \emph{output} of~|P| never gets stuck.

Predating us, \citet{Thiemann-combinators} and \citet{sumii-hybrid} deforested
the object term representation and expressed a partial evaluator as a collection
of term combinators in a typed metalanguage.  Like us, \citeauthor{sumii-hybrid}
follow \citet{Sperber-SelfApplicable} and \citet{asai-binding-time} in building
static and dynamic terms in tandem, to combine offline and online partial
evaluation.
\citearound{'s earlier self-reducer for the untyped
$\lambda$-calculus}\Citet{JFP-Mogensen} also
builds static and dynamic terms in tandem.  However, it
builds a static term for every object term, even a bound variable,
so it moves some work from |app| to~|pdyn| (in terms of Figure~\ref{fig:pe})
and remains untyped.  In contrast, we follow Sperber, Asai,
and \citeauthor{sumii-hybrid} in leaving the static term
optional, so as to perform lifting without
juggling explicit type indices in the encoding of an object term.

Our contribution to the literature on partial evaluation is to use mere
Hindley-Milner types in the metalanguage to assure statically and patently that
partially evaluating a well-typed object program not only never gets stuck but
also, if it terminates, produces a well-typed output program that never gets
stuck.  Moreover, these types form an instance of a general |Symantics|
signature that encompasses other interpreters such as evaluation and
compilation.  This early and manifest assurance of type safety contrasts, for
example, with \citearound{'s compiler generator (cogen) for
ML}\citet{Birkedal-PE-ML}, which transforms a program into its tagless
generating extension.  Because that cogen uses a universal type, the fact that
it never generates an ill-typed generating extension from a well-typed input
program is only manifest when each generating extension is type-checked, and the
fact that the generating extension never generates an ill-typed residual program
from well-typed static input is only manifest when each residual program is
type-checked.  Similarly, the fact that \citet{fiore:nbe-ppdp2002} and
\citets{balat:tdpe-popl2004} partial evaluator using delimited control
operators never turns well-typed code into ill-typed code is not assured by the
metalanguage, whether or not as part of a typed family of interpreter modules.
\begin{comment}
% Removed the following comment in view of Section 4.5 of Danvy's
% original TDPE paper (POPL 1996)
It is
type-directed, so the user must represent, as a term, the type of every
term to be partially evaluated.  We shift this work to the type checker
of the metalanguage.
By avoiding term-level type representations, our
approach makes it easier to perform algebraic simplifications (as
in~\S\ref{S:PE-solution}).
\end{comment}

Our partial evaluator reuses the
compiler~|C| and the evaluator~|R| by composing them.  This situation is
simpler than \citets{SperberThiemann:TwoForOne} composition of a partial
evaluator and a compiler, but the general ideas are similar.
\fi

@@

We would like to collect those interpretation functions into a module
with signature |Symantics|, to include the CBN CPS interpreter within our
general framework. Alas, as in~\S\ref{S:PE-problem}, the type of
an object term inductively determines the type of its interpretation:
the interpretation of an object term of type~$t$ may not have type
|(|$t$|->'w)->'w|, because $t$ may be a function type.  Again we
simulate a type function with a typecase distinction, using an extra
type argument to |repr|. Luckily, the type function |static| needed for
the partial evaluator 
in~\S\ref{S:PE-solution} is precisely the same type function we
need for CBN CPS\@.
\begin{code}
module RCN = struct
  type ('c,'sv,'dv) repr = {ko: 'w. ('sv -> 'w) -> 'w}
  let int (x:int) = {ko = fun k -> k x}
  let add e1 e2 = {ko = fun k ->
      e1.ko (fun v1 -> e2.ko (fun v2 -> k (v1 + v2)))}
  let if_ eb et ee = {ko = fun k ->
      eb.ko (fun vb -> if vb then (et ()).ko k else (ee ()).ko k)}
  let lam f = {ko = fun k -> k f}
  let app e1 e2 = {ko = fun k -> e1.ko (fun f -> (f e2).ko k)}
  let fix f = let rec fx f n = app (f (lam (fx f))) n in lam (fx f)
  let run x = x.ko (fun v -> v)
end
\end{code}

@@

The module returned by |CPST| does not define |repr|
and thus does not have signature |Symantics|.
The reason is again the type of |lam f|. Whereas
|int| and |add| return the (abbreviated) type
|('c, ..., (int -> 'w) -> 'w) S.repr|,
the type of \texttt{lam (add (int~1))} is
\begin{code}
('c, ..., ((int -> (int -> 'w1) -> 'w1) -> 'w2) -> 'w2) S.repr
\end{code}
Hence, to write the type equation defining |CPST.repr| we again need
a type function with a typecase distinction, similar to |static|
in~\S\ref{S:PE-solution}. Alas, the type function we need is not
identical to |static|, so we need to add another type argument to
|repr| in the |Symantics| signature. As in~\S\ref{S:PE-solution}, the
terms in previous implementations of |Symantics| stay unchanged, but the
|repr| type equations in those implementations have to take a new
(phantom) type argument.
The verbosity of these types is the only difficulty in defining a
new signature to replace |Symantics| that the output of |CPST| matches.

@@

Our partial evaluator establishes a bijection |static| between static
and dynamic types (the valid values of |'sv| and |'dv|), and between
static and dynamic terms.  It is customary to implement such a bijection
using an injection\hyp projection pair, as done for interpreters
\ifshort \citep{Ramsey-ML-module-mania,Benton-embedded-interpreters}\else
by \citet{Ramsey-ML-module-mania} and \citet{Benton-embedded-interpreters}\fi,
partial evaluation \ifshort \citep{Danvy-TDPE}\else by \citet{Danvy-TDPE}\fi,
and type-level functions \ifshort \citep{oliveira-typecase}\else by
\citet{oliveira-typecase}\fi.  As explained in~\S\ref{S:PE-solution}, we
avoid injection and projection at the type level by adding an argument
to |repr|.
Our solution could have been even more straightforward if MetaOCaml
provided total type-level functions such as |static| in
\S\ref{S:PE-solution}---simple type-level computations 
ought to become mainstream.

@@

@@

Our encoding of the type function |static| in \S\ref{S:PE-solution}
emulates type-indexed types and is related to intensional type analysis
\citep{Morrisett-intensional,Generic-Haskell}. However, our object
language and running examples in HOAS include |fix|,
which intensional type analysis cannot handle
\citep{xi-guarded}.
